Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-03-16 Paper/manuscript_v2/manuscript_07_01.tex
minor edits to the intro

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{bm}

@@

@@

\tableofcontents

@@

The key element of a GP is the covariance function that defines the dependence structure between function values at different inputs. However, the covariance function is also a computational issue because of the need of inverting its Gram matrix to optimize the hyperparameters. Given $n$ observations in the data, the computational complexity and memory requirements of exact GP implementation in general scale as $O(n^3)$ and $O(n^2)$, respectively. This limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of GPs \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \cite{hensman2017variational} presented a variational Fourier feature approximation for GPs that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random fourier features \citep{rahimi2008random,rahimi2009weighted}.
Further, certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline,Furrer+Nychka:2007}.
% , but not all spline models are GPs (e.g., some spline models are not marginalization consistent).

{\color{blue} Paul: Can we clarify how our discussed method relates to the former paragraph,
which introduces all sorts of basis function approximations. I.e. is our method
just one out of many or what makes it special? Perhaps we can somehow use the information provided
in: While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank. Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.}

Furthermore, a recent related work based on a spectral representation of GPs as an infinite series expansion with the Karhunen-Loève representation (see, e.g. \cite{grenander1981abstract}) can be found in \cite{JSSv090i10}.

In this paper we propose an approximate framework for fast and accurate inference for GPs. We focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Using basis function expansion, a GP is approximated with a linear model which makes inference considerably faster. A linear model structure makes GPs easier to implement as building blocks in more complicated models in modular probabilistic programming frameworks, where there is a big benefit if approximation specific computation is simple. Furthermore, a linear representation of a GP makes it easier to be used as latent function
in non-Gaussian observational models allowing modelling flexibility. The basis function approximation via Laplace eigenfunctions can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan}, WinBUGS \citep{lunn2000winbugs} and others. This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the relations going on among the key factors of the method such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the performance and accuracy of the method. Practical recommendations for the values of the key factors based on useful graphs that encode the recognized relationships are given. Which will help users to chose valid and optimized values for these factors, improving performance and saving time of computation. A diagnosis of whether the chosen values for the number of basis functions and the desired prediction space are adequate to fit to the actual data is proposed.

@@

The approach is implemented in a fully probabilistic framework and for the Stan programming probabilistic software. This work has served as a basis for the subsequent implementation of the method in the \textit{brms} library \citep{burkner2017brms} of the R software \citep{R2014R}. Several illustrative examples, simulated and real datasets, of the performance and applicability of the model, and accompanied by their Stan model codes, are developed.

Although there are several GP specific software packages available to date (GPML \citep{rasmussen2010gpml},  GPstuff \citep{vanhatalo2013gpstuff}, GPy \citep{gpy2014}, GPflow \citep{GPflow2017}), each provide efficient implementations only for a restricted range of GP based models. In this paper, we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks.

The remainder of the paper is structured as follows. In Section \ref{ch4_gp}, we introduce GPs with the covariance and spectral density functions. In Section \ref{ch5_sec_method}, the reduced rank approximation to GPs proposed by \cite{solin2018hilbert} is described. In Section \ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Sections \ref{ch5_sec_univariate_cases} and \ref{ch5_sec_multivariate_cases}. A brief conclusion of the work is made in Section \ref{ch5_sec_conclusion}. Appendix \ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind this Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function.

Online complementary material with more case studies illustrating the performance and applicability of the method can be found through the following link:
%
\begin{lstlisting}[breaklines]
https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/online_appendices
\end{lstlisting}

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}

@@

\author{TODO}

\date{}

@@

%\tableofcontents

@@

The key element of a GP is the covariance function that defines the dependence structure between function values at different inputs. However, evaluating the covariance function comes with a computational issue because of the need of inverting its Gram matrix to optimize the hyperparameters. Given $n$ observations in the data, the computational complexity and memory requirements of exact GP implementation in general scale as $O(n^3)$ and $O(n^2)$, respectively. This limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of GPs \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \cite{hensman2017variational} presented a variational Fourier feature approximation for GPs that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random fourier features \citep{rahimi2008random,rahimi2009weighted}.
Further, certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline,Furrer+Nychka:2007}.
Furthermore, a recent related work based on a spectral representation of GPs as an infinite series expansion with the Karhunen-Loève representation (see, e.g. \cite{grenander1981abstract}) can be found in \cite{JSSv090i10}.
% , but not all spline models are GPs (e.g., some spline models are not marginalization consistent).

%{\color{blue} Paul: Can we clarify how our discussed method relates to the former paragraph,
%which introduces all sorts of basis function approximations. I.e. is our method
%just one out of many or what makes it special? Perhaps we can somehow use the information provided
%in: While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank. Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.}

In this paper we propose and evaluate an new framework for fast and accurate inference for fully Bayesian GPs using Basis function approximations. We focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Using basis function expansion, a GP is approximated with a linear model which makes inference considerably faster. A linear model structure makes GPs easier to implement as building blocks in more complicated models in modular probabilistic programming frameworks, where there is a big benefit if approximation specific computation is simple. Furthermore, a linear representation of a GP makes it easier to be used as latent function
in non-Gaussian observational models allowing for more modelling flexibility. The basis function approximation via Laplace eigenfunctions can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan}, WinBUGS \citep{lunn2000winbugs} and others. This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to chose valid and optimized values for these factors, improving performance, and saving time of computation. A diagnosis of whether the chosen values for the number of basis functions and the desired prediction space are adequate to fit to the actual data is proposed.

@@

We have implemented the approach in the probabilistic programming language Stan \citep{carpenter2017stan} as well as subsequently in the \textit{brms} package \citep{burkner2017brms} of the R software \citep{R2019R}. Several illustrative examples of the performance and applicability of the method are shown using both simulated and real datasets. All examples are accompanied by the corresponding Stan and R code.

Although there are several GP specific software packages available to date, for example, GPML \citep{rasmussen2010gpml},  GPstuff \citep{vanhatalo2013gpstuff}, GPy \citep{gpy2014}, or GPflow \citep{GPflow2017}), each provide efficient implementations only for a restricted range of GP-based models. In this paper, we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks.

The remainder of the paper is structured as follows. In Section \ref{ch4_gp}, we introduce GPs with the covariance and spectral density functions. In Section \ref{ch5_sec_method}, the reduced rank approximation to GPs proposed by \cite{solin2018hilbert} is described. In Section \ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Sections \ref{ch5_sec_univariate_cases} and \ref{ch5_sec_multivariate_cases}. A brief conclusion of the work is made in Section \ref{ch5_sec_conclusion}. Appendix \ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind this Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. 
Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \texttt{Paper/online\_appendices}.
